---
title: "Haul-out behavior and detectability of bearded, ribbon, and spotted seals in the Bering and Chukchi Sea"
author:
  - Josh M. London
  - Paul B. Conn
  - Erin L. Richmond
  - Jay M. Ver Hoef
  - Stacie Hardy
  - Inrina S. Trukhanova
  - Michael F. Cameron
  - Peter L. Boveng
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    redoc::redoc
bibliography: references.bib
abstract: |
  Ice-associated seals rely on sea ice for a variety of activities, including breeding, molting, pupping, and rest.  In the Arctic, many of these activities occur in spring and early summer (April-June) as sea ice begins to melt and retreat northward.  Rapid acceleration of climate change in Arctic ecosystems is thus of concern, as the quantity and quality of suitable habitat is forecast to decrease.  In this paper, we use data collected from satellite-linked telemetry tags deployed between 2005 and 2017 to investigate the seasonal timing and environmental factors affecting sea ice use by phocid seals (specifically, bearded, ribbon, and spotted seals) in the Bering, Chukchi, and Beaufort Seas. In addition to providing baseline data on phenology, these data also allow us to refine availability estimates needed to accurately estimate abundance from aerial survey counts of seals basking on ice (i.e., to correct for the proportion of animals that are in the water while surveys are conducted).  Using generalized linear mixed pseudo-models to properly account for temporal autocorrelation, we fit models with a variety of covariates (e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to examine their ability to explain variation in hourly haul-out records.  We found evidence for strong diurnal and seasonal patterns in haul-out behavior, as well as strong weather effects (particularly wind and temperature).  In general, seals were more likely to bask on ice in the middle of the day and when wind speed was low and temperatures were higher.  Haul-out propensity increased through March and April, peaking in May and early June before declining again. The timing and frequency of haul-out events also varied substantially based on species and age-sex class.  For ribbon and spotted seals, models with year effects were highly supported, indicating that the timing and magnitude of haul-out behavior varied among years.  This analysis suggested a large degree of plasticity in the timing of annual haul-out peaks, suggesting that they may have ability to adapt breeding, molting, and pupping to environmental conditions.  Our analysis also emphasizes the importance of accounting for factors affecting haul-out behavior when interpreting the number of seals counted in aerial surveys. 
keywords: |
  availability; generalized linear mixed pseudo-model; haul-out behavior; phenology; Phocidae; sea ice
---


```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  include = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
```

```{r, libraries}
library(tidyverse)
library(redoc)
library(citr)

# library(buckshot) 
# Or use devtools::load_all('.', quiet = T)
```



# Introduction

Global climate change is causing considerable reductions in seasonal Arctic sea
ice extent (Comiso 2012). These reductions are a cause of concern for
ice-associated seals which depend on spring and early summer
sea ice (March-June) for diverse functions such as pupping, molting, breeding,
and rest (Boveng et al. 2009, Cameron et al. 2010, Kelly et al. 2010, Boveng et
al. 2013). There are several large knowledge gaps preventing biologists from
making definitive predictions about the ultimate effects of climate change on
the abundance and distribution of ice-associated seals in the Arctic. First,
knowledge about biological constraints on the phenology of reproductive and
molting behavior is generally lacking, so it is difficult to predict the
readiness with which ice-associated seal species will be able to adapt to future
changes (e.g., by adjusting pupping or molting schedules to earlier dates).
Second, trends in abundance of Arctic ice-associated seals are largely unknown,
so it is difficult to say what effect, if any, declines in sea ice habitat has
had on seal densities.

Ultimately, knowledge of trends (or lack thereof) in phenology and abundance (or
life history surrogates such as survival and recruitment) will be necessary to
make credible quantitative predictions of the effects of climate change on the
abundance and distribution of phocid seal populations. Before one can construct
a trend, however, one must first have credible data points. Several studies have
begun to estimate the distribution and abundance of ice-associated seal species
in the Arctic using aerial surveys (e.g., Bengtson et al. 2005, Conn et
al. 2014; Ver Hoef et al. 2014). Such studies must be conducted over large
areas, and estimation of absolute abundance requires making inference about
numerous nuisance processes affecting the observation of seals on ice, including
availability (only seals basking on ice are available to be counted), detection
probability (observers or automated detection systems may miss some seals on
ice), species misclassification, and possible disturbance of seals by aircraft
(Conn et al. 2014, Ver Hoef et al. 2014).

Quantitative studies on the phenology of pinniped sea ice use have been
conducted, but there are large knowledge gaps, especially relative to
ice-associated pinnipeds. Several authors have used logistic regression-style
analyses to estimate the probability of sea ice use from data garnered from
bio-loggers. In these models, haul-out
probabilities can be expressed as a function of predictive covariates, such as
time-of-day, day-of-year, sex, age class, and environmental conditions (e.g.,
Reder et al. 2003, Bengtson and Cameron 2004, Bengtson et al. 2005, Udevitz et
al. 2009, Ver Hoef et al. 2014). However, sample sizes have often been
underwhelming and insufficient to permit strong inference about life history
and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
Cameron’s (2004) study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.’s (2005) study was based on 6 telemetered ringed seals (Phoca
hispida) in the Chukchi Sea. These studies are often further limited by
logistical constraints on tagging fieldwork and the attachement duration
or operational life of
satellite tags. For ringed seals, tagging can often only occur after seals
emerge from snow dens in the spring (Bengtson and Cameron 2004). The use of head
mounted satellite tags provide some benefits over flipper mounted tags (e.g.
increased satellite transmittal rates), but these are lost during the annual
molt, limiting the effective length of haul-out timelines.

Despite these limitations, analyses of satellite telemetry records have proven
valuable for identifying covariates that explain variation in sea ice use by
ice-associated seals. For instance, in the Antarctic, Bengtson and Cameron
(2004) documented greater haulout propensity in tagged juvenile crabeater
(*Lobodon carcinophaga*) seals than adults, with highest probabilities in February
and at times close to solar noon. In the Arctic, Bengtson et al. (2005)
documented a higher propensity for sea ice use near solar noon, as did Ver Hoef
et al. (2014) in an analysis of bearded (*Erignathus barbatus*), ribbon
(*Histriophoca fasciata*), and spotted (*Phoca largha*) seals using much greater
sample sizes. Ver Hoef et al. (2014) also showed that haul-out probabilities
increased gradually starting in March, peaking in May and June for bearded,
ribbon, and spotted seals. However, their primary focus was estimating seal
abundance, and like a subsequent analysis by Conn et al. (2014), they did not
attempt to separately analyze haul-out schedules for different age- and
sex-classes (although initial models with sex and age as fixed effects were
insignificant at p = 0.05; Ver Hoef et al. 2014). Maximal haul-out probabilities
in these studies were near 0.65 for ringed seals (Bengtson et al. 2005),
and 0.66, 0.62, and 0.54 for bearded, ribbon, and spotted seals, respectively.

In this study, we use available ARGOS-linked bio-logger records in a
comprehensive investigation of sea ice use by ice-associated seals tagged in the
Bering and Chukchi Seas. Our goals are twofold. First, we wish to establish
baseline estimates of the chronology of sea ice use among different age- and
sex-classes. Second, we wish to refine estimates of haul-out availability
corrections for aerial surveys in order to improve estimates of seal abundance.
Previously estimated availability correction factors (e.g., Bengtson et
al. 2005, Conn et al. 2014, Ver Hoef et al. 2014) accounted for variables such
as the time-of-day and day-of-year, but did not investigate meteorological
variables that have been shown to influence haul-out behavior of walruses (Reder
et al. 2003, Udevitz et al. 2009). Further, previous correction factor analyses
assumed that tagged seals were representative of the population as a whole, even
though the age- and sex-distribution of telemetered individuals likely differs
considerably from the population. We thus seek to extend the scope of previous
haul-out analyses, investigating the influence of meteorological variability and
possible ramifications of nonrepresentative age-sex samples.

This article is structured as follows. First, we describe the bio-logging
data, including technical specifications of tags and information on tag
deployment. Second, we describe the modeling framework used to analyze telemetry
records, including details on explanatory covariates used to explain variation
in haul-out behavior. Third, we investigate differences in availability
probability when (i) the tagged sample is assumed to represent the population
(i.e. when age and sex are ignored), and (ii) when a stable stage distribution
is employed to scale inferences from tagged samples to the population. After
describing results, we discuss potential implications of our work.

# Methods

## Data collection

<!-- When and where (see also Fig. 1 and Tables 1-2)  -->
<!-- Tag details: Important differences between tags; longetivity of tags, etc. -->
<!-- Sensors, wet/dry   -->
<!-- Tag fixes (movement modeling?) -->

Tags that fall off due to molt or morality can still send data to satellites and
we did not want to include data from these tags in our analyses. As such, start
and end times of each tag deployment were made by experts who examined several
simultaneous data streams (e.g., tag locations and dive behavior) to determine
when tags seemed to be providing ecologically relevant information. Sensor
readings outside of these start and end times were discarded prior to analysis.

## Explanatory variables

In addition to sex and age class (“young-of-year,” sexually immature
“sub-adults”; and mature “adults”), we gathered a number of variables we
thought might help explain variation in haul-out probabilities (Table 2). These
included day-of-year (to help model seasonal effects) and solar hour (to help
model diurnal effects). Solar hour was calculated using the solaR package
(Perpinán 2012) within the R statistical environment (R Development Core
Team 2017) using modeled locations from satellite tags. We also compiled several
meteorological variables shown to affect haul-out behavior in other Arctic
pinnipeds (Reder et al. 2003, Udevitz et al. 2009). In particular, we linked
locations from satellite tags to weather reprojection values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction (Mesinger et al. 2006). The NARR model assimilates
observational data to produce a long-term picture of weather over North America.
For this study, NARR weather values were subset to the extent of our study area
over the Bering and Chukchi Seas at 3-hr intervals based on a grid resolution
of 32 km (1024 sq. km) The following meteorological variables were interpolated
and assigned to seal locations using a bilinear method: 1) air temperature at 2m
above the earth's surface, 2) wind consisting of northerly and easterly vector
components converted to wind speed using the Euclidean norm, 3) barometric
pressure at sea level, and 4) precipitation (Table 3). To assess whether the
peak of seal haul-out behavior was related to the amount of spring sea ice in
the Bering Sea, a metric for sea ice extent on April 1 was computed for each of
the years 2005-2016. We obtained sea ice concentration data from the Nimbus-7
SMMR and DMSP SSM/I-SSMIS Passive Microwave Dataset, Version 1 (Cavalieri et
al. 1996), and computed a sea ice index by summing the area of all Bering Sea
grid cells with sea ice concentration values greater than 0.15.

## Haul-out modeling

Haul-out records for seals are often characterized by multiple hours spent
basking on ice alternating with long periods of foraging (e.g. Fig. 2). Commonly
used statistical models for binary data (e.g. logistic regression) assume
independence among responses, an assumption that is clearly violated if hourly
responses are modeled. Any analysis that ignores temporal dependence in
responses will thus have overstated precision.

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs; Ver Hoef et
al. 2009) to model variation in haul-out behavior as a function of (1) covariate
predictors, (2) temporally autocorrelated random effects, and (3) individual
random effects representing heterogeneity in individual behavior. We used the
‘glmmLDTS’ package (Ver Hoef et al. 2010) to implement GLMPMs in the R
programming environment (R Development Core Team 2017). Owing to the large
number of records, we fit separate models to bearded, ribbon, and spotted seal
data sets. We fit two different models to our datasets. First, for each species,
we fitted a year-independent model that predicted average haul-out behavior as a
function of demographic, environmental, seasonal, and diurnal effects. Second,
for ribbon and spotted seals (which had considerably more data than bearded
seals), we fitted models that included all the effects from the first model, but
also permitted annual variation in haul-out timing. This second class of models
was used to examine whether haul-out patterns varied by year, and to determine
apparent peaks in haul-out behavior.
	
Models for all species include the following effects: age/sex class,
day-of-year, solar hour, temperature, wind speed, barometric pressure,
precipitation, and wind chill (represented by a wind × temperature interaction;
Udevitz et al. 2009), together with day-of-year and time-of-day interactions to
permit diurnal patterns to change throughout the year (information on these
variables are provided in Table 3). Spotted and ribbon seal models included
interactions between day-of-year and age-sex class, but we omitted these from
bearded seal models due to poor representation of age-sex classes (Table 1).
However, bearded seal models included a latitudinal effect (and an interaction
with day-of-year) since bearded seals occupy a substantial range and we were
interested in possible differences in the timing of haul-out along a latitudinal
gradient. We omitted the latitudinal effect from ribbon and spotted seal models
since these species are typically most prevalent near the southern ice edge in
the Bering Sea (Conn et al. 2014) and such an effect would be less interpretable
in this case.

We followed Ver Hoef et al. (2014) in using linear, quadratic, and cubic effects
of day-of-year to represent seasonal changes in behavior. However, unlike
previous models for harbor seals (Ver Hoef et al. 2009) and ice-associated seals
(Ver Hoef et al. 2014), which treated hour-of-day as a 24-level categorical
variable to capture diurnal cycles, we adopted a continuous formulation based on
Fourier series that provides a flexible model while preserving the inherit
circularity needed for time-of-day effects (i.e., hour 0 should be equal to
hour 24). It also represents hour-of-day with 6 variables, which is a
considerable reduction when compared to a 24 variable model, especially when
fitting models with interactions between hour-of-day and other variables (e.g.,
age-sex class, day-of-year). According to this approach, we used the following
specification for hour-of-day effects:

H\_t=∑\_(i=1)^3▒〖α\_i1 cosa(πt/2(i+1) ) 〗+α\_i2 sina(πt/2(i+1) ),

where H\_t gives the effect for solar hour t and α\_ij for i = 1, 2, or 3 and j
= 1 or 2 are estimated parameters (regression coefficients).

For the second set of models examining inter-annual variation in sea ice use, we
fitted models with year by day-of-year interactions. However, in this case we
only included year*day and year*day2, omitting the main effects of year as well
as year\*day3 interactions as models with the latter effects were numerically
unstable. However, the modeled interactions were sufficient to allow shifts in
haul-out distribution, as one can show mathematically that a simple horizontal
shift in timing of haul-out distributions does not affect the main effects or
cubic terms in a polynomial regression model.

After fitting GLMPM models, we used “type III” F-tests to calculate p-values for
candidate predictors (Ver Hoef et al. 2009). We also produced predictions of
haul-out behavior as a function of influential predictors. For models with
inter-annual variation in haul-out distributions, we based predictions on
spatially-averaged weather covariate values over the Bering Sea. Such
predictions were then used to determine annual peaks in haul-out activity. We
then used standard linear models in R (function “lm”) to regress the day these
peak occurred on April sea ice extent in the Bering Sea.

## Stable stage distributions

Previous availability correction factors used in aerial surveys for
ice-associated Arctic seals (e.g. Bengtson et al. 2005, Conn et al. 2014, Ver
Hoef et al. 2014) ignored any age- and sex-based variation in haul-out
probabilities. The implicit assumption is that the behavior of telemetered
animals represents the population as a whole. However, if haul-out probabilities
are related to sex- and age-class, and if there are tangible differences between
the age- and sex-structure of the telemetered sample and that of the population,
correction factors computed in this way may lead to biased abundance estimates.

To determine how much of a potential problem this is for our data, we compared
estimates of haul-out probabilities that (1) ignored sex- and age-structure, and
(2) accounted for sex- and age-structure, adjusting for the relative proportion
of seals in each sex- and age-class. To simplify comparisons, we fitted two
additional GLMPMs to ribbon and spotted seal haul-out datasets. In the first
model, we expressed haul-out probabilities as a function of solar hour and
day-of-year. In the second, we also included effects of sex-age class and
sex-age interactions with solar hour and day-of-year. To adjust for the
proportion of seals in each age class, we calculated a predicted
population-averaged haul-out probability at solar noon as a function of
day-of-year (t) as

(p_t ) ̃=∑_s▒∑_a▒〖ω_(s,a) p ̂_(s,a,t) 〗 ,
	
where ω\_(s,a) is the proportion of the population of seals in age class a and
sex s. Typically, we do not know ω\_(s,a) exactly, but can use information about
likely mortality and reproductive schedules of seals to approximate ω\_(s,a)
using stable stage distributions calculated from matrix population models
(Caswell 2001).

To calculate plausible values for ω\_(s,a) we obtained natural mortality
predictions from a hierarchical meta-analysis of phocid natural mortality
(Trukhanova et al. 2018). We then assumed a 50/50 sex ratio at birth, 1 pup per
litter, and reproductive schedules as reported by Fedoseev (2000) for seals in
the western Bering Sea to formulate Leslie matrices Ai for each species i.
Stable age distributions v\_i={v\_i0,v\_i1,⋯,v\_iA} were then determined by
normalizing the dominant eigenvector of Ai so that they summed to one
(Caswell 2001). Given a stable age distribution, we calculated the proportion of
the population by sex and stage (young-of-year: 0-1 year; subadult: at least one
year but sexually immature; adult: sexually mature) using data on sexual
maturity from Fedoseev (2000). In particular, the expected proportion of the
population that are sex s and young-of-year was set as π\_(i,s,yoy)=0.5v\_i0,
while the expected proportion that are subadults and adults were set to

π\_(i,s,sub)=〖0.5∑\_a▒〖〖(1-m〗\_(i,s,a))〗 v〗\_ia and

π\_(i,s,ad)=〖0.5∑\_a▒m\_(i,s,a) v〗\_ia, respectively.

Here, m\_(i,s,a) gives the anticipated proportion of animals of species i and
sex s that are sexually mature. Further details on mortality, reproductive and
maturity schedules, and matrix modeling methods are presented in Appendix A.

## Computing

We compiled a standalone R package, ‘buckshot,’ which includes all data and R
scripts needed to reproduce analyses. It is currently available on github at
<https://github.com/jmlondon/buckshot> and will be published and archived at a
long-term, publically available repository upon manuscript acceptance.

# Results

## Haul-out modeling

Models omitting year effects suggested that day-of-year, solar hour, age-sex
class, temperature, and wind substantially altered haul-out behavior of all
three species, with F tests producing p-values less than 0.05 for variables
embodying these effects and/or their interactions (Appendix B). Haul-out
probabilities typically increased throughout March and April, reaching a peak in
May and early June before declining again. Diurnal patterns were present, with
maximum haul-out probabilities shortly after solar noon. Spotted seals appeared
to initiate longer bouts of haul-out behavior before ribbon seals and to have a
longer period of such behavior. Both ribbon and spotted seals had much larger
probabilities of basking on ice than bearded seals.

The probability of hauling out on ice decreased with wind speed (Fig. 4), and
increased with temperature (Fig. 5). Bearded and ribbon seals also appeared to
haul out more when barometric pressure was high, while spotted seals appeared to
haul out less as precipitation increased (Appendix B-C). Bearded seals farther
north initiated haul-out activity later in the spring, maintaining this behavior
for longer periods (Appendix B, Fig. 6). Wind chill (as embodied by a
temperature by wind interaction) substantially decreased bearded seal haul-out
probabilities but did not appear to be statistically significant for spotted or
ribbon seals. (Appendix B).

The second set of models that included annual variation in haul-out patterns
produced p-values that were strongly significant for linear (ribbon seals:
F10,76662=9.92, p\<0.0001; spotted seals: F8,51981=5.23, p\<0.0001) and
quadratic (ribbon seals: F10,76662=8.02, p\<0.0001; spotted seals:
F8,51981=3.79, p=0.0002) year-by-day interactions. Predicted distributions of
haul-out activity were largely unimodal, but varied substantially among and
within years with respect to both the timing and magnitude of haul-out peaks
(Figs 7-8). For instance, in 2010 and 2016 spotted seal haul-out activity peaked
in early May, while in other years (e.g. 2011) it appeared to peak three weeks
later. It is important to note that predicted variation in annual haul-out
patterns likely reflects both process error and sampling variability. For
instance, some of the more extreme patterns (e.g. ribbon seals in 2012 , spotted
seals in 2014) were for years with only one tagged seal and are unlikely to
represent population-level processes.

The timing of haul-out peaks did not appear to be related to the total
concentration of sea ice in the Bering Sea on April 1 of each year for either
species. In particular, linear models produced effects of sea ice that were
close to zero and p-values substantially larger than 0.05 (ribbon: F1,8=0.21,
p=0.66; spotted F1,6=1.46, p=0.27).

## Stable stage distributions

Stable stage distributions calculated using mortality and reproductive schedules
produced different patterns for the four species of ice-associated seals
residing in Alaskan waters (Fig. 9). Ribbon seals mature the fastest, and so can
be expected to have the largest proportion of adults and smallest proportion of
subadults; conversely, ringed seals mature slowly and are predicted to have
roughly equal proportions of adults and subadults (note that we do not attempt
to model ringed seal haul-out behavior in this study; see Discussion).
Population-level haul-out distributions were somewhat different depending on
whether we a) adjusted haul-out probabilities based on stable stage
distributions or b) assumed the tagged sample represented the population. In
particular, ribbon seal estimates differed somewhat at the beginning of the time
series, but the two approaches produced similar estimates (Fig. 10) within the
window that previous aerial surveys in the region have been conducted (April and
May; e.g. Conn et al. 2014). By contrast, during this time period spotted seal
haul-out estimates were often ≈10% higher for the approach that adjusted for
stable age distributions, presumably because subadults had lower predicted
haul-out probabilities than adults but made up a disproportionately high
proportion of the tagged sample.

# Discussion

In this paper, we fit statistical models to data from sensors on satellite
tagged ice-associated seals in order to examine factors affecting the
probability that Arctic seals haul out on ice. Such probabilities are important
for understanding the timing and ecology of seal behavior, since haul out peaks
are associated with important life history functions such as breeding, pupping
and molting. Knowledge of factors affecting haul-out probabilities is a first
step in being able to predict possible impacts of losses of sea ice on
ice-associated seals. Understanding the factors affecting haul-out probabilities
is also necessary to properly interpret seal counts made during aerial surveys.
For instance, a count might be low because there are few seals, or because the
timing and conditions are such that many seals are in the water.

Our analysis revealed that seals use sea ice progressively more throughout the
spring, and with peak use near mid-May to early June before declining again. We
also found that seals preferentially haul out on ice shortly after solar noon,
which may be associated with fewer prey in the water column (??? Can someone
else who knows more about their feeding ecology contribute here?).
Interestingly, bearded seals appeared to have two peaks in haul-out activity
throughout the day, one shortly after solar noon, and one centered near solar
midnight (related to benthic feeding?). It appeared that adult spotted seals
initiated haul-out behavior earlier in the year than ribbon seals, and
maintained this behavior over a longer duration; however they exhibited stronger
diel behavior than ribbon seals (Fig 3). These patterns are similar to those
obtained by Conn et al. (2014), who used a subset of the data we analyzed here.

Unlike previous analyses of seal haul-out data (e.g. Ver Hoef et al. 2009, Conn
et al. 2014), we also investigated the influence of sex-age class on haul-out
probabilities. Although it was difficult to discern a pattern for bearded seals
owing to low sample size, it appeared that adult male spotted and ribbon seals
typically hauled out for longer durations, perhaps because they were less
constrained by energetic demands associated with nursing and could sustain
longer periods on ice to engage in molting and mating activities. Subadult
ribbon and spotted seals exhibited different behavior when compared to adults.
In particular, subadult ribbon seals exhibited similar haul-out behavior to
adults, whereas subadult spotted seals appeared to haul-out less frequently and
sustained a shorter duration of peak haul-out behavior than adults. The latter
result is likely due to subadults only needing to haul out on ice to molt;
pupping and mating are not yet in their behavioral repertoire.

We also investigated the influence of environmental covariates on haul-out
probabilities, including wind speed, temperature, barometric pressure,
precipitation, and wind chill. These have been investigated for walrus (e.g.
Udevitz et al. 2009), but to our knowledge have never been investigated for
ice-associated seals. In particular, we found clear evidence that seal haul-out
probabilities decrease with wind speed and increase with temperature (Figs 4-5).
In addition, we found evidence that ribbon seal haul-out probabilities are
positively related to barometric pressure, and that spotted seal haul-out
probabilities are negatively related to precipitation (Appendix C). The tendency
of seals to haul out in “good” weather may be related to energetic
considerations, better predator detection, or both. Bearded seal haul-out
behavior was predicted to change slightly based on how far north tagged seals
were; seals farther north tended to initiate haul-out behavior slightly later,
and for a longer period of time (Fig. 6). This response may be related to
different temperature regimes, or to timing of ice melt, which occurs later in
locations farther north.

There appeared to be rather substantial annual deviations in the timing and
magnitude of haul-out peaks (Figs 7-8). For example, peak ribbon seal haul-out
behavior appeared to vary by as much as three weeks between seasons. There thus
seems to be a reasonable amount of plasticity in the timing of haul-out,
suggesting some plasticity in life history functions such as pupping, molting
and breeding. However, we were unable to relate the timing of haul-out
conditions to underlying ice conditions, which makes predicting climate change
consequences on life history events difficult. Future investigation into
environmental factors contributing to the timing of haul-out peaks seems
warranted. For instance, Ray et al. (2010) suggested that ribbon and spotted
seals were most dependent on loose pack ice at the sourthern sea ice edge; thus,
the relative suitability for haul-out behavior may have more to do with the
characterization, position, and quantity of loose pack ice rather than the total
amount of ice in the Bering Sea. Further work to characterize remotely sensed
sea ice data into relevant ice types (e.g. loose pack, continuous pack, rounded
pack) would be useful.

Previous attempts to estimate the abundance of phocid seals from aerial survey
data in the Bering and Chukchi Seas (e.g. Bengtson et al. 2005, Conn et
al. 2014, Ver Hoef et al. 2014) have used estimated haul-out probabilities to
correct for the proportion of animals that are in the water and thus unavailable
to be counted during aerial surveys. Although several of these studies allowed
haul-out probabilities to vary by day-of-year and time-of-day, previous
correction factors have not accounted for variability between years, different
weather conditions, and the age-sex class of the sample. In this paper, we have
shown that there can be considerable differences in the number of seals hauled
out on ice based on these factors. We thus recommend that future abundance
analyses employ correction factors that account for these factors. For instance,
it is relatively straightforward to obtain weather reanalysis products for times
and locations that are surveyed and to construct a relevant correction factor
based on predictions of GMPMs. Further, one can apply stable stage distribution
estimates from matrix population models to adjust predictions to the “average”
prediction of the population. The latter suggestion is particularly appropriate
when the age structure of the tagged sample does not reflect the likely age
structure of the population. For instance, population weighted estimates can
differ substantially from those based on a naïve analysis of the tagged sample,
as was the case with analysis of spotted seal data (e.g. Fig 9). The most
challenging element in developing availability correction factors is with annual
variability. It can be difficult to get a sufficient sample size to estimate
year-specific correction factors, particularly because research teams would
likely need to tag seals and conduct aerial surveys concurrently which requires
considerably more personnel and money. One possible suggestion is to estimate a
“shift” parameter within models for aerial survey counts that allow the peak
of haul-out distributions to be adjusted earlier or later in the year based on
the frequency of counts observed over time. Regardless, researchers should
anticipate there being some unmodeled heterogeneity in availability probability
present in abundance estimates obtained from aerial surveys. This may make trend
detection difficult, as one will not know if moderate differences in abundance
estimates are attributable to changes in abundance or changes in haul-out
frequencies.

Predictions of absolute haul-out probability in this paper were somewhat
different than those previously reported for these species, especially for
bearded seals. For instance, Ver Hoef et al. (2014) and Conn et al. (2015) used
haul-out correction factors with maximums of 0.66 for bearded seals, 0.62 for
ribbon seals, and 0.54 for spotted seals, where maximums corresponded to times
near solar noon in mid-late May. Applying models that ignore age, sex, and year
effects, these probabilities were 0.38, 0.72, and 0.60, respectively, under the
current analysis framework. Our current estimates reflect increased sample sizes
in terms of number of animals, but also improvements to the way data are
prepared prior to analysis. For instance, for bearded seals there were often
long stretches at the end of the tag history where it appeared animals were
hauled out 100% of the time. In reality, these likely represented tags shed
during molt on ice floes or washed up on beaches. In the present study, experts
used multiple data streams to determine effective start and end dates for each
deployment, and these long stretches were often removed. In previous studies,
such periods were often retained and likely led to positively biased estimates
of haul-out probability. Given the potential ramifications for abundance
estimation, we strongly suggest conducting such data screening exercises prior
to analyzing haul-out data. Future research should be devoted to automating this
procedure, perhaps using artificial intelligence algorithms to classify tag
behavior based on multiple data streams. Such a framework would require adequate
training data, e.g. using tags purposefully placed on ice floes, beaches, or
subject to tag fouling so that automatic classifiers could appropriately
discriminate properly functioning tags from anomalous behavior. Another approach
would be to integrate additional information on pulse or respiration, as can now
be done with life history tags (e.g., Horning and Hill 2005). At minimum, we
suggest reporting haul-out correction factors used when conducting abundance
analyses so that attendant estimates can be revisited as our state of knowledge
about tag function and haul-out behavior evolves.

We focused this paper on haul-out behavior of bearded, ribbon, and spotted
seals. Ringed seals are also present in the Bering and Chukchi Seas but exhibit
qualitatively different haul-out behavior. For instance, many ringed seals build
subnivean layers in the snow on top of the sea ice, which they frequently
inhabit before snow melt causes their lairs to collapse (Kelly 1988). Thus, the
wet-dry sensor on a satellite tag could indicate that an animal is hauled out,
but it could still be unavailable to be detected because it is in a lair. We
hope to address availability of ringed seals using data from satellite tags,
replicate survey tracks, and auxiliary information about snow depth and timing
of melt in a future study.

# Acknowledgements

The findings and conclusions in the paper are those of the author(s) and do not
necessarily represent the views of the National Marine Fisheries Service, NOAA.
Any use of trade, product, or firm names does not imply an endorsement by the
U.S. Government. Funding for this study was provided by the U.S. National
Oceanic and Atmospheric Administration. MMPA permit \#?

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
git2r::repository(here::here())
```
